### 2024-08-01 paper/main.tex
final cleanup

\newpage
\doublespace %
\linenumbers %
\SectionHeader{ABSTRACT}
\\
Indecisive behaviour can be catastrophic, leading to car crashes or stock market losses. Despite fruitful efforts across several decades to understand decision-making, there has been little research on what leads to indecision. Here we examined how indecisions arise under high-pressure deadlines. In our first experiment participants attempted to select a target by either reacting to a stimulus or guessing, when acting under a high pressure time constraint. We found that participants were suboptimal, displaying a below chance win percentage due to an excessive number of indecisions. Computational modeling suggested that participants were excessively indecisive because they failed to account for a time delay and temporal uncertainty when switching from reacting to guessing, a phenomenon previously unreported in the literature. In a followup experiment we provide direct evidence for functionally relevant time delay and temporal uncertainty when switching from reacting to guessing. Collectively, our results demonstrate that humans are suboptimal and fail to account for a time delay and temporal uncertainty when switching from reacting to guessing, leading to indecisive behaviour.

@@

\begin{figure}[H]
  \centering
  \includegraphics[scale =1]{figures/exp1_experimental_design_v2.png}
  \caption*{\boldblue{Figure 1: Experimental Design.} \boldblue{A)} Participants grasped the handle of a robotic manipulandum and made reaching movements in the horizontal plane. An LCD projected images (start position, targets) onto a semi-silvered mirror. Participants began each trial by moving their cursor (purple) into the start position (solid white circle). At the start of the trial (0 ms), they heard a tone and saw two targets (white rings) and a timing bar (white rectangle) appear on the screen. During each trial, a stimulus (yellow cursor) would move quickly in a straight line to one of the two targets. Participants were instructed to reach the same target as the stimulus within a time constraint of 1500ms. This time constraint was visually represented with a timing bar (white rectangle) that decreased in width according to the elapsed time. \boldblue{B)} A trial was considered a win and the participant received one point if they successfully reached the same target as the stimulus within the time constraint. A trial was considered incorrect and the participant received zero points if they reached the opposite target as the stimulus within the time constraint. A trial was considered an indecision and the participant received 0 points if they failed to reach a target within the time constraint. \boldblue{C)} Stimulus movement onset on each trial was randomly drawn from a specific probability distribution in each condition. Using a within experimental design, we manipulated the mean (early, middle, late) and standard deviation (low, high) of the stimulus movement onset probability distribution. For the main manuscript we focus on results for the low variance conditions, with high variance conditions results shown in \boldblue{Supplementary A.}}
\end{figure}

@@

\noindent Participant movement onsets for the low variance conditions are shown in \boldblue{Fig. 2A}, We found a significant main effect of stimulus movement onset mean (F[1.55,29.48] = 4.36, p = 0.030) and variance (F[1.00,19.00], p = 0.017). There was no significant interaction between stimulus movement onset mean and variance (F[1.66,31.47], p = 0.565). When collapsed across low and high variance, participant movement onsets were significantly greater in the \emph{middle mean} conditions compared to the \emph{early mean} conditions (p = 0.014, $\hat{\theta}$ = 72.5 \%), suggesting that participants waited longer to react to the stimulus movement and guessed later in time. Again when collapsed across low and high variance, participant movement onset significantly decreased from the \emph{middle mean} conditions to the \emph{late mean} conditions (p = 0.018, $\hat{\theta}$ = 62.5\%). Here, an earlier participant movement onset in the \emph{late mean} condition suggests that participants attempted to wait and react to the stimulus, but ended up guessing.

Participant movement onset standard deviation for the low variance conditions are shown in \boldblue{Fig. 2B}. There was a main effect of mean (F[1.38, 26.28], p = 0.018) and variance (F[1.00,19.00], p < 0.001) of the stimulus movement onset, and no significant interaction (F[1.98, 37.58], p = 0.097).

@@

      \includegraphics[scale = 1]{figures/exp1_data_panel_with_models.png}
  \end{minipage}\hfill
  \begin{minipage}[c][\pdfpageheight][t]{0.5\textwidth}
      \caption*{
        \boldblue{Figure 2: Behaviour Results}. \boldblue{A)} Participant movement onset, \boldblue{B)} participant movement onset standard deviation, \boldblue{C)} indecisions, \boldblue{D)} wins, and \boldblue{E)} incorrects are shown for each condition. In our high time pressure task, participants made a high proportion of indecisions. Open circles are individual participant data. Filled icons are the predicted behavior of the three models, see legend for detail. Interestingly, we also found that the average win percentage (41.25\%) was significantly below the 50\% chance level (p < 0.001) in the \emph{late mean} condition, clearly demonstrating suboptimal behaviour. Critically, participants would have earned more points if they had simply guessed earlier on all trials, rather than attempting to react to the stimulus. Model Results: Only the Partial Switch Time Model, which had only a partial representation of the time delay and uncertainty when switching from reacting to guessing, was able to predict the indecisions and a suboptimal win rate (below 50\%) in the \emph{late mean} condition. See \boldblue{Fig. 3} for the time varying model predictions for the \emph{late mean} condition. Collectively, our findings show that humans are suboptimal and excessively indecisive.} 
  \end{minipage}
\end{figure}

@@

The win percentage across all conditions was 56.25\% (range: 6.2\%, 93.8\%; \boldblue{Fig. 2D}). We found a significant interaction between stimulus movement onset mean and variance for wins (F[1.54, 29.30] = 23.73, p<0.001). The \emph{late mean} condition had significantly less wins than the \emph{early mean} condition (p < 0.001, $\hat{\theta}$ = 95.0\%). Interestingly, in the \emph{late mean} condition we found that the average win percentage was significantly below the 50\% chance level (p<0.001; $\hat{\theta}$ = 95.0\%), which was true for 19 out of 20 participants. Since guessing on every trial would lead to a win percentage of 50\%, the only way participants would be below chance is if they are excessively indecisive.
The incorrect percentage across all conditions was 26.3\% [range: 0.0\%, 57.5\%; \boldblue{Fig. 2E}]. We found significant interactions between stimulus movement onset mean and variance for incorrect decisions (F[1.66,31.51]=3.72, p=0.033). Participants displayed a greater percentage of incorrect decisions in the \emph{late mean} condition than the \emph{early mean} condition (p < 0.001, $\hat{\theta}$ = 92.5\%).

@@

\noindent Our behavioural findings in \boldblue{Experiment 1} demonstrated that participants were suboptimal decision makers. Through our modelling efforts, we were able to capture this suboptimal decision-making by including a switch time delay and uncertainty when transitioning from reacting to guessing. The switch time delay and uncertainty were only partially represented by the Partial Switch Time Model when determining the optimal time to switch from reacting and guessing. However, we are not aware of any work that considers the temporal time delays and uncertainty associated with switching from reacting to guessing within a trial. The goal of \boldblue{Experiment 2} was to determine if there is indeed a switch delay and uncertainty that occurs when humans transition from reacting to guessing.

@@

  \caption*{\boldblue{Figure 4: Experiment 2 Design.} The goal of this experiment was to test the idea that there is a delay and uncertainty associated with switching from reacting to guessing, as suggested by our findings in  Experiment 1. \boldblue{A)} Participants responded to two different types of stimuli. In the react trials (pink), the stimulus (yellow cursor) would move to one of the two potential targets. Participants were instructed to reach the same target as the stimulus as quickly as possible. \boldblue{B)} In the guess trials (blue), the stimulus disappeared from the start position. Once the stimulus disappeared, participants were instructed to guess which target the stimulus would appear in and select that target as quickly as possible. After the participant reached the target, the stimulus’s cursor would appear in one of the targets. \boldblue{C)} We had three experimental conditions. In the \emph{react or guess} condition, react trials and guess trials were randomly interleaved (50 react trials and 50 guess trials). Participants were informed that the stimulus would either move to one of the targets or disappear. In the \emph{only react} condition, participants were informed that the stimulus would alway move to one of the two targets (50 react trials and 0 guess trials). They were also told that the stimulus would not disappear. In the \emph{only guess} condition, participants were informed that the stimulus would always disappear (0 react trials and 50 guess trials). They were also informed the stimulus would not move. }
\end{figure}

@@

In \boldblue{Experiment 1}, participants were required to reach the same target as a cursor before a time constraint. Within a trial, they could either react to a moving stimulus or guess which of the two targets would be correct. We saw that 95\% of participants had a win rate less than chance (50\%) in the \emph{late mean} condition, which corresponded with an average of 19.4\% indecisions. This proportion of indecisions aligns with our recent prior work that examined competitive human-human decision-making with a high time pressure\autocite{lokeshHumansUtilizeSensory2022}. In this competitive task, one participant attempted to reach the same target as their opponent, while the other tried to reach the opposite target within a time constraint. It was suggested that the high proportion of indecisions in this competitive human-human task were the result of participants waiting too long to acquire sensory information of their opponent, despite the impending time deadline. Likewise, we found a high proportion of indecisions across experimental conditions. Our results would also suggest that participants waited to acquire sensory information of when the stimuli would move. Moreover, building upon Lokesh and colleagues (2022), our work suggests that a key contributor leading to excessive indecisions is failing to account for the time delay and temporal uncertainty when switching from reacting to guessing.

Past work has suggested that humans can nearly optimally account for time delays and temporal uncertainty when performing decision-making \autocite*{balciRiskAssessmentMan2009,jazayeriTemporalContextCalibrates2010,balciOptimalTemporalRisk2011} and movement tasks \autocite*{hudsonOptimalCompensationTemporal2008,deanTradingSpeedAccuracy2007} when attempting to maximize reward. Here we considered two optimal models, the No Switch Time Model and Full Switch Time Model, which both had full knowledge of all available time delays and temporal uncertainties. Interestingly, both the No Switch Time Model and Full Switch Time Model showed that even when fully accounting for all sensorimotor delays and uncertainties, indecisions were a part of an optimal strategy in all but one of the six conditions. In other words, given the inherent delays and uncertainty of our nervous system\autocite{faisalNoiseNervousSystem2008}, an optimal strategy of earning maximal reward may involve indecisive behaviour on some proportion of trials. We are unaware of any work in the literature suggesting that some level of indecisions may be optimal. Even though making some indecisions can be optimal, our results in \boldblue{Experiment 1} were in support of the idea that humans are suboptimal. Specifically, in the \emph{late mean} condition, we found that humans were suboptimal since they had a win percentage lower than chance, which arose from an excessive number of indecisions. The Partial Switch Time Model was suboptimal, since it had a partial representation of time delays and temporal uncertainties associated with switching from reacting to guessing. We found that this model best explained behaviour, including a below chance win percentage and an excessive number of indecisions. The Partial Switch Time Model supports the notion that humans suboptimally select decision times when under high time pressures. An interesting future direction would be to test whether different reward structures, such as placing a higher reward on wins or punishing indecisions\autocite*{kahnemanProspectTheoryAnalysis2013,rothReinforcementbasedProcessesActively2023,rothPunishmentLeadsGreater2024,galeaDissociableEffectsPunishment2015}, would provide a means to reduce an excessive number of indecisions.

The combined empirical evidence of \boldblue{Experiment 1} and computational modelling suggested the existence of time delay and uncertainty when switching from reacting to guessing. However, we were unaware of any work in the literature to support this idea. In \boldblue{Experiment 2} we tested the notion of a more delayed and uncertain response time when switching from reacting to guessing, compared to guessing by itself. Indeed, we found that when participants had to switch from reacting to guessing, their response times were significantly slower and more uncertain than when they only had to guess. One possibility for increased time delays and temporal uncertainty could be related to switching between different processing ‘modes’. In our task, participants may have switched from a `react mode’ that corresponded to preparing to follow the stimulus, and then switch to a `guess mode’ to randomly select one of the targets.

Dutilh and colleagues (2011) explored the idea of switching between a stimulus controlled (i.e. react) mode and a guess mode between trials \autocite{dutilhPhaseTransitionModel2011}. In their task, participants were required to discriminate between a word stimuli from a non-word stimuli by selecting one of two buttons during a two-alternative forced choice task. Between trials, the authors manipulated whether participants received more reward for fast decisions to promote guessing, or more reward for accurate decisions that promoted reacting. Participants displayed longer response times when transitioning from more accurate decisions to fast decisions, compared to when transitioning from fast decisions to accurate decisions given the same current reward weighting. The authors interpreted these results to represent a resistance, termed hysteresis, when switching between react and guess modes. That is, participants are more likely to stay in their current mode than switch modes. They also highlighted that classical decision-making models, such as drift-diffusion models \autocite{ratcliffModeling2alternativeForcedchoice2018} or more recently the urgency-gating model \autocite*{cisekDecisionsChangingConditions2009,derosiereTradingAccuracySpeed2021,thuraDecisionMakingUrgency2012} do not consider different modes of reacting or guessing. Extending upon the findings of Dutilh and colleagues (2011) that examined between trial mode switching, our work suggests that humans switch between react and guess modes within a trial. Importantly, we find that not accounting for the time delays and temporal uncertainties when switching from reacting to guessing gives rise to excessive indecisions. To our knowledge, it is unknown how different modes would be represented in the nervous system. One possibility is that the different modes represent different attractors from a neural dynamical systems perspective \autocite*{erlhagenDynamicFieldTheory2002,wangDecisionMakingRecurrent2008,churchlandNeuralPopulationDynamics2012,shenoyCorticalControlArm2013}, which would be an interesting avenue of investigation.

We found in \boldblue{Experiment 2} that participant response times were more delayed by approximately 75ms during guess trials in the \emph{react or guess} condition, compared to guess only trials. Moreover, the response time difference between guess and react trials in the \emph{react or guess} condition are significantly greater than the response time difference between the guess only and react only conditions. Collectively, these results suggest that participants have an initial preference to react, before having to switch to a guess in the \emph{react or guess} condition. In this experiment behaviour may be explained by a strong preference to react since it yields a 100\% success probability, as opposed to guessing that on average yields a 50\% success probability. It would be interesting for future studies to examine if they can manipulate the magnitude or probability of reward to switch a preference between reacting or guessing, and how this impacts indecisive behaviour.

@@

Here we showed in our first experiment humans are excessively indecisive under time constraints. Computational work and second experiment suggest that indecisive behaviour can occur by not accounting for the time delays and temporal uncertainty associated with switching from reacting to guessing. Our experimental and theoretical approach offers a new paradigm to study indecisions, which has received surprisingly little attention despite its ecological relevance. This work advances how indecisive behaviour arises, which is important to understand when attempting to avoid potentially catastrophic events during high time pressure scenarios.

@@

\vspace{2mm}
\noindent\boldblue{\large Experiment 1 Design}

@@

\noindent The goal of \boldblue{Experiment 2} was to test whether there is a time delay and uncertainty associated with switching from reacting to guessing. To investigate, participants began each trial by moving their cursor into a start position. Then, a stimulus cursor appeared in the start position. Each trial began with a beep and the two potential targets appeared on the screen. Here participants experienced two different types of trials: react trials or guess trials. The react trials consisted of the stimulus moving to one of the two targets (\boldblue{Fig. 4A}). Participants were instructed to follow the stimulus as quickly as possible. The guess trials consisted of the stimulus disappearing from the start position (\boldblue{Fig. 4B}). Participants were instructed to select the target they believed the stimulus would end up in as quickly as possible. After the participant selected their target, the stimulus cursor appeared in one of the two targets. The stimulus movement onset (reaction trials) or disappearance time (guess trials) was drawn from a normal distribution (mean = 800 ms, standard deviation = 50 ms).  The stimulus randomly selected the left and right targets with equal probability.

@@

\noindent \emph{Participant movement onset}: On each trial, we found when the time point where the participant hand velocity exceeded 0.05 m/s \autocite{gribbleRoleCocontractionArm2003,calaloSensorimotorSystemModulates2023}. The meantime point across all trials within a condition was used to estimate participant movement onset.

>>>>

\newpage
\doublespace %
\linenumbers %
\SectionHeader{ABSTRACT}
\\
Indecisive behaviour can be catastrophic, leading to car crashes or stock market losses. Despite fruitful efforts across several decades to understand decision-making, there has been little research on what leads to indecision. Here we examined how indecisions arise under high-pressure deadlines. In our first experiment participants attempted to select a target by either reacting to a stimulus or guessing, when acting under a high pressure time constraint. We found that participants were suboptimal, displaying a below chance win percentage due to an excessive number of indecisions. Computational modeling suggested that participants were excessively indecisive because they failed to account for a time delay and temporal uncertainty when switching from reacting to guessing, a phenomenon previously unreported in the literature. In a followup experiment we provide direct evidence for a functionally relevant time delay and temporal uncertainty when switching from reacting to guessing. Collectively, our results demonstrate that humans are suboptimal and fail to account for a time delay and temporal uncertainty when switching from reacting to guessing, leading to indecisive behaviour.

@@

\begin{figure}[H]
  \centering
  \includegraphics[scale =1]{figures/exp1_experimental_design_v2.png}
  \caption*{\boldblue{Figure 1: Experimental Design.} \boldblue{A)} Participants grasped the handle of a robotic manipulandum and made reaching movements in the horizontal plane. An LCD projected images (start position, targets) onto a semi-silvered mirror. Participants began each trial by moving their cursor (purple) into the start position (solid white circle). At the start of the trial (0 ms), they heard a tone and saw two targets (white rings) and a timing bar (white rectangle) appear on the screen. During each trial, a stimulus (yellow cursor) would move quickly in a straight line to one of the two targets. Participants were instructed to reach the same target as the stimulus within a time constraint of 1500 ms. This time constraint was visually represented with a timing bar (white rectangle) that decreased in width according to the elapsed time. \boldblue{B)} A trial was considered a win and the participant received one point if they successfully reached the same target as the stimulus within the time constraint. A trial was considered incorrect and the participant received zero points if they reached the opposite target as the stimulus within the time constraint. A trial was considered an indecision and the participant received 0 points if they failed to reach a target within the time constraint. \boldblue{C)} Stimulus movement onset on each trial was randomly drawn from a specific probability distribution in each condition. Using a within experimental design, we manipulated the mean (early, middle, late) and standard deviation (low, high) of the stimulus movement onset probability distribution. For the main manuscript we focus on results for the low variance conditions, with high variance conditions results shown in \boldblue{Supplementary A.}}
\end{figure}

@@

\noindent Participant movement onset for the low variance conditions is shown in \boldblue{Fig. 2A}. We found a significant main effect of stimulus movement onset mean (F[1.55,29.48] = 4.36, p = 0.030) and variance (F[1.00,19.00], p = 0.017). There was no significant interaction between stimulus movement onset mean and variance (F[1.66,31.47], p = 0.565). When collapsed across low and high variance, participant movement onsets were significantly greater in the \emph{middle mean} conditions compared to the \emph{early mean} conditions (p = 0.014, $\hat{\theta}$ = 72.5\%), suggesting that participants waited longer to react to the stimulus movement and guessed later in time. Again, when collapsed across low and high variance, participant movement onset significantly decreased from the \emph{middle mean} conditions to the \emph{late mean} conditions (p = 0.018, $\hat{\theta}$ = 62.5\%). Here, an earlier participant movement onset in the \emph{late mean} condition suggests that participants attempted to wait and react to the stimulus, but ended up guessing.

Participant movement onset standard deviation for the low variance conditions is shown in \boldblue{Fig. 2B}. There was a main effect of mean (F[1.38, 26.28], p = 0.018) and variance (F[1.00,19.00], p < 0.001) of the stimulus movement onset, and no significant interaction (F[1.98, 37.58], p = 0.097).

@@

      \includegraphics[scale = 1]{figures/exp1_data_panel_with_models.png}
  \end{minipage}\hfill
  \begin{minipage}[c][\pdfpageheight][t]{0.5\textwidth}
      \caption*{
        \boldblue{Figure 2: Behaviour Results}. \boldblue{A)} Participant movement onset, \boldblue{B)} participant movement onset standard deviation, \boldblue{C)} indecisions, \boldblue{D)} wins, and \boldblue{E)} incorrects are shown for each condition. In our high time pressure task, participants made a high proportion of indecisions. Open circles are individual participant data. Filled icons are the predicted behavior of the three models, see legend for detail. Interestingly, we also found that the average win percentage (41.25\%) was significantly below the 50\% chance level (p < 0.001) in the \emph{late mean} condition, clearly demonstrating suboptimal behaviour. Critically, participants would have earned more points if they had simply guessed earlier on all trials, rather than attempting to react to the stimulus. \boldblue{Model Results}: Only the Partial Switch Time Model, which had only a partial representation of the time delay and uncertainty when switching from reacting to guessing, was able to predict the indecisions and a suboptimal win rate (below 50\%) in the \emph{late mean} condition. See \boldblue{Fig. 3} for the time varying model predictions for the \emph{late mean} condition. Collectively, our findings show that humans are suboptimal and excessively indecisive.} 
  \end{minipage}
\end{figure}

@@

The win percentage across all conditions was 56.25\% (range: 6.2\%, 93.8\%; \boldblue{Fig. 2D}). We found a significant interaction between stimulus movement onset mean and variance for wins (F[1.54, 29.30] = 23.73, p<0.001). The \emph{late mean} condition had significantly less wins than the \emph{early mean} condition (p < 0.001, $\hat{\theta}$ = 95.0\%). Interestingly, in the \emph{late mean} condition we found that the average win percentage was significantly below the 50\% chance level (p<0.001; $\hat{\theta}$ = 95.0\%), which was true for 19 out of 20 participants. Since guessing on every trial would lead to a win percentage of 50\%, the only way participants would be below chance is if they were excessively indecisive.
The incorrect percentage across all conditions was 26.3\% [range: 0.0\%, 57.5\%; \boldblue{Fig. 2E}]. We found significant interactions between stimulus movement onset mean and variance for incorrect decisions (F[1.66,31.51] = 3.72, p = 0.033). Participants displayed a greater percentage of incorrect decisions in the \emph{late mean} condition than the \emph{early mean} condition (p < 0.001, $\hat{\theta}$ = 92.5\%).

@@

\noindent Our behavioural findings in \boldblue{Experiment 1} demonstrated that participants were suboptimal decision makers. Through our modelling efforts, we were able to capture this suboptimal decision-making by including a switch time delay and uncertainty when transitioning from reacting to guessing. The switch time delay and uncertainty were only partially represented by the Partial Switch Time Model when determining the optimal time to switch from reacting and guessing. However, we are not aware of any work that considers a delay and uncertainty associated with switching from reacting to guessing within a trial. The goal of \boldblue{Experiment 2} was to determine if there is indeed a switch delay and uncertainty that occurs when humans transition from reacting to guessing.

@@

  \caption*{\boldblue{Figure 4: Experiment 2 Design.} The goal of this experiment was to test the idea that there is a delay and uncertainty associated with switching from reacting to guessing, as suggested by our findings in  Experiment 1. \boldblue{A)} Participants responded to two different types of stimuli. In the react trials (pink), the stimulus (yellow cursor) would move to one of the two potential targets. Participants were instructed to reach the same target as the stimulus as quickly as possible. \boldblue{B)} In the guess trials (blue), the stimulus disappeared from the start position. Once the stimulus disappeared, participants were instructed to guess which target the stimulus would appear in and select that target as quickly as possible. After the participant reached the target, the stimulus cursor would appear in one of the targets. \boldblue{C)} We had three experimental conditions. In the \emph{react or guess} condition, react trials and guess trials were randomly interleaved (50 react trials and 50 guess trials). Participants were informed that the stimulus would either move to one of the targets or disappear. In the \emph{only react} condition, participants were informed that the stimulus would alway move to one of the two targets (50 react trials and 0 guess trials). They were also told that the stimulus would not disappear. In the \emph{only guess} condition, participants were informed that the stimulus would always disappear (0 react trials and 50 guess trials). They were also informed the stimulus would not move. }
\end{figure}

@@

In \boldblue{Experiment 1}, participants were required to reach the same target as a cursor before a time constraint. Within a trial, they could either react to a moving stimulus or guess which of the two targets would be correct. We saw that 95\% of participants had a win rate less than chance (50\%) in the \emph{late mean} condition, which corresponded with an average of 19.4\% indecisions. This proportion of indecisions aligns with our recent prior work that examined competitive human-human decision-making with a high time pressure\autocite{lokeshHumansUtilizeSensory2022}. In this competitive task, one participant attempted to reach the same target as their opponent, while the other tried to reach the opposite target within a time constraint. It was suggested that the high proportion of indecisions were the result of participants waiting too long to acquire sensory information of their opponent, despite the impending time deadline. Likewise, we found a high proportion of indecisions across experimental conditions. Our results would also suggest that participants waited to acquire sensory information of when the stimuli would move. Moreover, building upon Lokesh and colleagues (2022), our work suggests that a key contributor leading to excessive indecisions is failing to account for the time delay and temporal uncertainty when switching from reacting to guessing.

Past work has suggested that humans can nearly optimally account for time delays and temporal uncertainty when performing decision-making \autocite*{balciRiskAssessmentMan2009,jazayeriTemporalContextCalibrates2010,balciOptimalTemporalRisk2011} and movement tasks \autocite*{hudsonOptimalCompensationTemporal2008,deanTradingSpeedAccuracy2007} when attempting to maximize reward. Here we considered two optimal models, the No Switch Time Model and Full Switch Time Model, which both had full knowledge of all available time delays and temporal uncertainties. Interestingly, both the No Switch Time Model and Full Switch Time Model showed that even when fully accounting for all sensorimotor delays and uncertainties, indecisions were a part of an optimal strategy in all but one of the six conditions. In other words, given the inherent delays and uncertainty of our nervous system\autocite{faisalNoiseNervousSystem2008}, an optimal strategy of earning maximal reward may involve indecisive behaviour on some proportion of trials. We are unaware of any work in the literature suggesting that some level of indecisions may be optimal. Even though making some indecisions can be optimal, our results in \boldblue{Experiment 1} were in support of the idea that humans are suboptimal. Specifically, in the \emph{late mean} condition, we found that humans were suboptimal since they had a win percentage lower than chance, which arose from an excessive number of indecisions. The Partial Switch Time Model was suboptimal, since it had a partial representation of the time delay and temporal uncertainties associated with switching from reacting to guessing. We found that this model best explained behaviour, including a below chance win percentage and an excessive number of indecisions. The Partial Switch Time Model supports the notion that humans suboptimally select decision times when under high time pressures. An interesting future direction would be to test whether different reward structures, such as placing a higher reward on wins or punishing indecisions\autocite*{kahnemanProspectTheoryAnalysis2013,rothReinforcementbasedProcessesActively2023,rothPunishmentLeadsGreater2024,galeaDissociableEffectsPunishment2015}, would provide a means to reduce an excessive number of indecisions.

The combined empirical evidence of \boldblue{Experiment 1} and computational modelling suggested the existence of a time delay and uncertainty when switching from reacting to guessing. However, we were unaware of any work in the literature to support this idea. In \boldblue{Experiment 2} we tested the notion of a more delayed and uncertain response time when switching from reacting to guessing, compared to guessing by itself. Indeed, we found that when participants had to switch from reacting to guessing, their response times were significantly slower and more uncertain than when they only had to guess. One possibility for increased time delays and temporal uncertainty could be related to switching between different processing ‘modes’. In our task, participants may have switched from a `react mode’ that corresponded to preparing to follow the stimulus, to a `guess mode’ to randomly select one of the targets.

Dutilh and colleagues (2011) explored the idea of switching between a stimulus controlled (i.e. react) mode and a guess mode between trials \autocite{dutilhPhaseTransitionModel2011}. In their task, participants were required to discriminate between a word stimuli from a non-word stimuli by selecting one of two buttons during a two-alternative forced choice task. Between trials, the authors manipulated whether participants received more reward for accurate decisions to promote reacting, or more reward for fast decisions that promoted guessing. Participants displayed longer response times when transitioning from more accurate decisions to fast decisions, compared to when transitioning from fast decisions to accurate decisions given the same current reward weighting. The authors interpreted these results to represent a resistance, termed hysteresis, when switching between react and guess modes. That is, participants are more likely to stay in their current mode than switch modes. They also highlighted that classical decision-making models, such as drift-diffusion models \autocite{ratcliffModeling2alternativeForcedchoice2018} or more recently the urgency-gating model \autocite*{cisekDecisionsChangingConditions2009,derosiereTradingAccuracySpeed2021,thuraDecisionMakingUrgency2012}, do not consider different modes of reacting or guessing. Extending upon the findings of Dutilh and colleagues (2011) that examined between trial mode switching, our work suggests that humans switch react and guess modes within a trial. Importantly, we find that not accounting for the time delays and temporal uncertainties when switching from reacting to guessing gives rise to excessive indecisions. To our knowledge, it is unknown how different modes would be represented in the nervous system. One possibility is that the different modes represent different attractors from a neural dynamical systems perspective \autocite*{erlhagenDynamicFieldTheory2002,wangDecisionMakingRecurrent2008,churchlandNeuralPopulationDynamics2012,shenoyCorticalControlArm2013}, which would be an interesting avenue of investigation.

We found in \boldblue{Experiment 2} that participant response times were more delayed by approximately 75 ms during guess trials in the \emph{react or guess} condition, compared to guess only trials. Moreover, the response time difference between guess and react trials in the \emph{react or guess} condition are significantly greater than the response time difference between the \emph{guess only} and \emph{react only} conditions. Collectively, these results suggest that participants have an initial preference to react, before having to switch to a guess. In this experiment behaviour may be explained by a strong preference to react since it yields a 100\% success probability, as opposed to guessing that on average yields a 50\% success probability. It would be interesting for future studies to examine if they can manipulate the magnitude or probability of reward to switch a preference between reacting or guessing, and how this impacts indecisive behaviour.

@@

Here we showed in our first experiment that humans are excessively indecisive under time constraints. Computational work and a second experiment suggested that indecisive behaviour can occur by not accounting for the time delay and temporal uncertainty associated with switching from reacting to guessing. Our experimental and theoretical approach offers a new paradigm to study indecisions, which has received surprisingly little attention despite its ecological relevance. This work advances how indecisive behaviour arises, which is important to understand when attempting to avoid potentially catastrophic events during high time pressure scenarios.

@@

\vspace{5mm}
\noindent\boldblue{\large Experiment 1 Design}

@@

\noindent The goal of \boldblue{Experiment 2} was to test whether there is a time delay and uncertainty when switching from reacting to guessing. To investigate, participants began each trial by moving their cursor into a start position. Then, a stimulus cursor appeared in the start position. Each trial began with a beep and the two potential targets appeared on the screen. Here participants experienced two different types of trials: react trials or guess trials. The react trials consisted of the stimulus moving to one of the two targets (\boldblue{Fig. 4A}). Participants were instructed to follow the stimulus as quickly as possible. The guess trials consisted of the stimulus disappearing from the start position (\boldblue{Fig. 4B}). Participants were instructed to select the target they believed the stimulus would end up in as quickly as possible. After the participant selected their target, the stimulus cursor appeared in one of the two targets. The stimulus movement onset (reaction trials) or disappearance time (guess trials) was drawn from a normal distribution (mean = 800 ms, standard deviation = 50 ms).  The stimulus randomly selected the left and right targets with equal probability.

@@

\noindent \emph{Participant movement onset}: On each trial, we found when the time point where the participant hand velocity exceeded 0.05 m/s \autocite{gribbleRoleCocontractionArm2003,calaloSensorimotorSystemModulates2023}. The mean time point across all trials within a condition was used to estimate participant movement onset.
